The dynamics of risk perception in a Mediterranean agroecosystem

### 2020-02-12 manuscript/manuscript.Rmd
adjust drought length

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library(raster)
library(ncdf4)
library(tidyverse)
set.seed(1000)
```

@@

# Drought risks in eastern Mediterranean {#sec:2}

@@

Risk Perception
In order to manage risk, decision-makers must be perceive what exactly that risk is in the local environment.
An individual's perceived risk entails more than simply the "objective" risk in the external environment [@Tucker2013].
A person's perceived risk depends in large part on their past experience, and how those experiences shape their expectations for the future.
Likewise, the social responses to risk will refelct the distribution of perceived risks in a population, and by extension the collective memory of past droughts will have consequences for the population's survival and evolution.

Belief-based agents
We can capture the influences of risk perception computationally by simulating a population of "belief-based agents".
These agents develop beliefs about the risk of drought in their environment through personal experience. 
Each agent's perceived drought risks will change their subjective expected crop yields from planting different ratios of wheat to barley.

Bayesian brains
We can efficiently simulate these population-level risk perceptions in a population of agents with Bayesian brains (Kahvalti et al 2019 double check). Bayesian learning allows for a minimal representation of the environemnt, so that each agent's beliefs can be encoded in two parameters -- defining the Beta distribution over their prior beliefs -- rather than the record of all observed years. Not only does this approach have a clear computational efficiency, but for the same reason this reflects the processes actually going on in the heads of decision makers.
Our brains do not record every bit of perceived informaiton in memory, rather it stores a "compact encoding" of that information which it uses for future decisions. Even if individuals aren't -- consciously or unconsciously --- making these calculations -- the basic algorithmic problem faced by the brain, and solutions it has evolved, reflect the same basic sets of constraints (Sanborn and Chater 2016).

Bayesian learning from bernoulli observations.
We can treat this problem elegantly using bayesian statistics using online bayesian learning from bernoulli objersbations (Bissiri 2010). Put simply, the probability of a drought occuring in any given year is treated as a coin flip (i.e. bernoulli distribution) with paramter theta reprenting the risk of drought. Uncertainty in the value of theta can be represented with a beta distribution, constrained between 0 and 1 (Figure XX). We can thus represent the exact information content of each individual agent's subjective experience of droughts using the diffusion of the prior belief. What we see here is how an indiviudal's perception of the risk of drought may start off diffuse at first (here as an uniformative flat prior in which all values of theta are equally likely. As each year of experience is observed, the agent updates its beliefs accordingly, balancing the information in this year's observation with the cumulative weight of experience.

@@

But what if nature isn't stationary?
That is, what if the risk of a drought, theta, is not a constant but rather varies unpredictably in time?
The statistical problem shifts then to one of change detection (o'reilly 2013, gallistel et al 2014).
In other words do people simply adapt to changes -- as in the fixed bayesian learner and moving average approaches -- or do they detect them?
Strict year-to-year updating can result in overly conservative risk perceptions, suggesting people are slow to change their beliefs when in reality they often quickly update them (Bissiri 2010).
Indeed, the "forgetful bayesian brain" model seems to perform better, with a dynamic rather than fixed bayesian belief model (Zhang and Yu 2013).

@@

# The Model
```{r}
agents <- tibble(id = 1:150, 
                 alpha = 1.5,
                 beta = 1.5,
                 risk = alpha / (alpha + beta),
                 age = sample.int(80, size = 150, replace = TRUE),
                 crop = 'barley')
```

```{r}
learn1 <- function(agents, drought) {
  if (drought == 'TRUE') wheat_yield <- wheat_dry; barley_yield <- barley_dry
  if (drought == 'FALSE') wheat_yield <- wheat_normal; barley_yield <- barley_normal

  agents %>%
    # farmer's turn to decide what to plant
      mutate(yield_exp = expected_yield(risk),
         barley_exp = expected_yield_barley(risk),
         wheat_exp = expected_yield_wheat(risk),
         expected_yield = barley_exp + wheat_exp,
         crop = case_when(barley_exp > wheat_exp ~ 'barley',
                              barley_exp < wheat_exp ~ 'wheat',
                              barley_exp == wheat_exp ~ 'either'),
         prediction = map2(alpha, beta, ~rbeta(100, .x, .y)), # sample 100 as if agent is thinking in percentages
         wheat_prop = map_dbl(prediction, ~sum(. < indiff$risk)) / 100) %>%
      select(-prediction) %>%
  # nature's turn to decide realized yields
    mutate(payoff = wheat_yield * wheat_prop + barley_yield * (1 - wheat_prop),
           alpha = drought + alpha,
           beta = beta + 1 - drought,
          # mu = alpha / (alpha + beta),
           risk = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}

# or keep risk parameter constant and reduce the effective sample size, in other words keep the ratio of alpha to beta but decrease the absoulte value
```

@@

This suggests that people are more in danger of not realizing conditions are improving when they actually have -- the opportunity cost -- rather than failing to realize the change in the firstplace.
The law of small numbers is very important here. The range of each person's experience is so small -- every individual's perceived risk will be a small and noisy sample of the true value.
Older people are much more certain, even though they are wrong.
```{r}

droughts <-   c(runif(40) < 0.2, runif(20) < 0.4, runif(40) < 0.2)
```

```{r}
simulation1 <- accumulate(droughts, ~learn1(.x, .y), .init = agents) %>% 
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time)) %>%
  filter(time != 1)
```

Redo these plots so its a single person, one who was born in a drought period then experiences an ameliorarion and another who experiences a shift to drought conditions later in life? basically this section neds to work more with individuals and explore that stuff

```{r, fig.height = 6, fig.width = 6}
test1 <- tibble(time = 1:100,
       exp1 = rep(0.2, 100),
       exp2 = rep(0.4, 100),
       exp3 = c(rep(0.2, )))

ggplot(simulation1, aes(time, risk)) +
  #geom_line(data = test1) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  geom_smooth(color = 'darkorange') +
  theme_classic()
```

```{r}
ggplot(simulation2, aes(time, risk)) +
  geom_line(data = test1) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic()

```

```{r}
ggplot(simulation2, aes(time, risk)) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic()
```

```{r}
simulation1 %>%
  group_by(time, crop) %>%
  tally() %>%
ggplot(aes(time, n, fill = crop)) +
  geom_bar(width = 1, stat = 'identity', position = 'fill') +
  scale_fill_manual(values = c('darkgoldenrod', 'lightgoldenrod')) +
  theme_classic()
```

```{r}
simulation2 %>%
  group_by(time, crop) %>%
  tally() %>%
ggplot(aes(time, n, fill = crop)) +
  geom_bar(width = 1, stat = 'identity', position = 'fill') +
  scale_fill_manual(values = c('darkgoldenrod', 'lightgoldenrod')) +
  theme_classic()
```

@@

Note in these plots above and below we're summing all the wheat and barley in the population, compared to below which shows the wheat proportions of individuals
```{r}
ggplot(simulation1, aes(time, wheat_prop)) +
  geom_point(alpha = .3) +
  theme_classic()

ggplot(simulation2, aes(time, wheat_prop)) +
  geom_point(alpha = .3) +
  theme_classic()
```

```{r}
simulation1 %>%
  group_by(time) %>%
  summarise(famine = sum(payoff == 0)) %>%
ggplot(aes(time, famine)) +
  geom_line() +
  theme_classic()
simulation2 %>%
  group_by(time) %>%
  summarise(famine = sum(payoff == 0, na.rm = TRUE)) %>%
ggplot(aes(time, famine)) +
  geom_line() +
  theme_classic()
```

```{r}
simulation1 %>%
  group_by(time) %>%
  summarise(payoff = mean(payoff)) %>%
ggplot(aes(time,payoff)) +
  geom_point(aes(color = time)) +
  theme_classic()

simulation2 %>%
  group_by(time) %>%
  summarise(payoff = mean(payoff, na.rm = TRUE)) %>%
ggplot(aes(time,payoff)) +
  geom_point(aes(color = time)) +
  theme_classic()
```

```{r}
library(ggridges)
simulation1 %>%
  mutate(time_step = 5 * round(time/5),
         famine = payoff == 0) %>%
  ggplot(aes(wheat_prop, as.factor(time_step))) +
  geom_density_ridges(rel_min_height = 0.005) +
  theme_ridges()

simulation2 %>%
  mutate(time_step = 5 * round(time/5),
         famine = payoff == 0) %>%
  ggplot(aes(wheat_prop, as.factor(time_step))) +
  geom_density_ridges(rel_min_height = 0.005) +
  theme_ridges()
```

@@

\section{Results and Discussion}

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library(raster)
library(ncdf4)
library(tidyverse)
```

@@

# Holocene drought risks in the eastern Mediterranean {#sec:2}

@@

In order to manage risk, decision-makers must be perceive what exactly that risk is in the local environment.
An individual's perceived risk entails more than simply the "objective" risk in the external environment [@Tucker2013].
A person's perceived risk depends in large part on their past experience, and how those experiences shape their expectations for the future.
Likewise, the social responses to risk will refelct the distribution of perceived risks in a population, and by extension the collective memory of past droughts will have consequences for the population's survival and evolution.

We can capture the influences of risk perception computationally by simulating a population of "belief-based agents."
These agents develop beliefs about the risk of drought in their environment through personal experience. 
Each agent's perceived drought risks will change their subjective expected crop yields from planting different ratios of wheat to barley.

## Bayesian brains
We can efficiently simulate these population-level risk perceptions in a population of agents with Bayesian brains (Kahvalti et al 2019 double check). Bayesian learning allows for a minimal representation of the environemnt, so that each agent's beliefs can be encoded in two parameters -- defining the Beta distribution over their prior beliefs -- rather than the record of all observed years. Not only does this approach have a clear computational efficiency, but for the same reason this reflects the processes actually going on in the heads of decision makers.
Our brains do not record every bit of perceived informaiton in memory, rather it stores a "compact encoding" of that information which it uses for future decisions. Even if individuals aren't -- consciously or unconsciously --- making these calculations -- the basic algorithmic problem faced by the brain, and solutions it has evolved, reflect the same basic sets of constraints (Sanborn and Chater 2016).

### Bayesian learning from bernoulli observations
We can treat this problem elegantly using bayesian statistics using online bayesian learning from bernoulli objersbations (Bissiri 2010). Put simply, the probability of a drought occuring in any given year is treated as a coin flip (i.e. bernoulli distribution) with paramter theta reprenting the risk of drought. Uncertainty in the value of theta can be represented with a beta distribution, constrained between 0 and 1 (Figure XX). We can thus represent the exact information content of each individual agent's subjective experience of droughts using the diffusion of the prior belief. What we see here is how an indiviudal's perception of the risk of drought may start off diffuse at first (here as an uniformative flat prior in which all values of theta are equally likely. As each year of experience is observed, the agent updates its beliefs accordingly, balancing the information in this year's observation with the cumulative weight of experience.

@@

### The problem of nonstationarity
That is, what if the risk of a drought, theta, is not a constant but rather varies unpredictably in time?
The statistical problem shifts then to one of change detection (o'reilly 2013, gallistel et al 2014).
In other words do people simply adapt to changes -- as in the fixed bayesian learner and moving average approaches -- or do they detect them?
Strict year-to-year updating can result in overly conservative risk perceptions, suggesting people are slow to change their beliefs when in reality they often quickly update them (Bissiri 2010).
Indeed, the "forgetful bayesian brain" model seems to perform better, with a dynamic rather than fixed bayesian belief model (Zhang and Yu 2013).

@@

# Risk perception in a Mediterranean agroecosystem

So what are the consequences of these cognitive processes for early farmers in the Mediterranean.

```{r}
agents <- tibble(id = 1:150, 
                 alpha = 1.5,
                 beta = 1.5,
                 risk = alpha / (alpha + beta),
                 #age = sample.int(80, size = 150, replace = TRUE),
                 crop = 'barley')
```

```{r}
learn <- function(agents, drought) {
  if (drought == 'TRUE') wheat_yield <- wheat_dry; barley_yield <- barley_dry
  if (drought == 'FALSE') wheat_yield <- wheat_normal; barley_yield <- barley_normal

  agents %>%
    # farmer's turn to decide what to plant
      mutate(yield_exp = expected_yield(risk),
         barley_exp = expected_yield_barley(risk),
         wheat_exp = expected_yield_wheat(risk),
         expected_yield = barley_exp + wheat_exp,
         crop = case_when(barley_exp > wheat_exp ~ 'barley',
                              barley_exp < wheat_exp ~ 'wheat',
                              barley_exp == wheat_exp ~ 'either'),
         prediction = map2(alpha, beta, ~rbeta(100, .x, .y)), # sample 100 as if agent is thinking in percentages
         wheat_prop = map_dbl(prediction, ~sum(. < indiff$risk)) / 100) %>%
      select(-prediction) %>%
  # nature's turn to decide realized yields
    mutate(payoff = wheat_yield * wheat_prop + barley_yield * (1 - wheat_prop),
           alpha = drought + alpha,
           beta = beta + 1 - drought,
          # mu = alpha / (alpha + beta),
           risk = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}
```

@@

This suggests that people are more in danger of not realizing conditions are improving when they actually have -- the opportunity cost -- rather than failing to realize the change in the firstplace.
The law of small numbers is very important here. The range of each person's experience is so small -- every individual's perceived risk will be a small and noisy sample of the true value.
Older people are much more certain, even though they are wrong.

```{r}
make_droughts <- function() c(runif(25) < 0.2, runif(25) < 0.4, runif(25) < 0.2)

set.seed(11111)
droughts <- make_droughts()
```

```{r}
run_sim <- function(droughts) {
  accumulate(droughts, ~learn(.x, .y), .init = agents) %>% 
    bind_rows(.id = 'time') %>%
    mutate(time = as.numeric(time)) %>%
    filter(time > 1)
}

sim1 <- run_sim(droughts)
```

```{r, fig.height = 4, fig.width = 6}
test1 <- tibble(time = 1:75,
       risk = c(rep(0.2, 25), rep(0.4, 25), rep(0.2, 25)))

ggplot(sim1, aes(time, risk)) +
  geom_line(data = test1) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
 # geom_smooth(color = 'darkorange') +
  theme_classic()
```

```{r}
sim1 %>%
  group_by(time, crop) %>%
  tally() %>%
ggplot(aes(time, n, fill = crop)) +
  geom_bar(width = 1, stat = 'identity', position = 'fill') +
  scale_fill_manual(values = c('darkgoldenrod', 'lightgoldenrod')) +
  theme_classic()
```

```{r}
library(furrr)
plan(multisession)

sim2 <- tibble(rep = 1:500,
       droughts = replicate(500, make_droughts(), simplify = FALSE)) %>%
  mutate(results = future_map(droughts, run_sim)) %>%
  select(-droughts) %>%
  unnest(results)
```

@@

Note in these plots above and below we're summing all the wheat and barley in the population, compared to below which shows the wheat proportions of individuals.

```{r}
sim2 %>%
  group_by(rep, time) %>%
  summarise(test = mean(payoff))%>%
  summarise(test = mean(test)) %>%
  #summarise(famine = sum(payoff == 0) / 150) %>%
  #summarise(famine = mean(famine)) %>%
  ggplot(aes(test)) +
  geom_histogram()
#ggplot(aes(time, famine)) +
#  geom_line() +
#  theme_classic()

sim2 %>%
  group_by(rep, time) %>%
  summarise(test = mean(payoff))%>%
  ggplot(aes(time, test, color = rep)) +
  geom_point(alpha = .5)
```

```{r}
sim1 %>%
  group_by(time) %>%
  summarise(payoff = mean(payoff)) %>%
ggplot(aes(time,payoff)) +
  geom_point(aes(color = time)) +
  theme_classic()
```

```{r}
library(ggridges)
sim1 %>%
  mutate(time_step = 5 * round(time/5),
         famine = payoff == 0) %>%
  ggplot(aes(wheat_prop, as.factor(time_step))) +
  geom_density_ridges(rel_min_height = 0.005) +
  theme_ridges()
```

@@

# Discussion and Conclusions
